---
title: "Heat Map Generating"
output: html_document
---

```{r setup, include=FALSE}
library(terra)
library(sf)
library(dplyr)
library(stringr)
library(purrr)
library(exactextractr)
library(spatstat)
library(raster)

terraOptions(tempdir = tempdir())
``` 

# 1. Load Input Data

```{r load-data}
# Load actual data paths
output_dir <- "C:/Users/yolan/OneDrive/Documenten/UGENT/Master/Stage/R/aandachtsoorten/"
crss = " +proj=lcc +lat_0=90 +lon_0=4.36748666666667 +lat_1=49.8333339 +lat_2=51.1666672333333 +x_0=150000.01256 +y_0=5400088.4378 +ellps=intl +units=m +no_defs  "

cell_size <- 5 #default celgroote voor dit project

chm_ras1 = raster('wh_vh1m.tif', crs = crss)
chm_ras1[chm_ras1 < 0] <- NA

reference_raster = aggregate(chm_ras1, fun = mean, fact = cell_size/res(chm_ras1))
species_points <- st_read("species_points_processed.gpkg")




plant_polygons <- st_read("plant_vlakken_processed.shp")
```

# 2. Resample Raster to Match 1m Resolution

```{r resample-raster}
# Make sure both layers have the same CRS as the reference raster
species_points <- st_transform(species_points, crs(reference_raster))
plant_polygons <- st_transform(plant_polygons, crs(reference_raster))

```

# 3. Filter Points and Polygons by Year and Type

```{r filter-points-polygons}
filtered_species_points <- species_points %>%
  mutate(JAAR = as.numeric(JAAR)) %>%
  filter(JAAR > 2014, TYPE == "Aandachtssoort")

filtered_plant_polygons <- plant_polygons %>%
  mutate(JAAR = as.numeric(JAAR)) %>%
  filter(JAAR > 2014)

```

# 4. Remove Overlapping Polygons per Species

```{r remove-overlapping}
plant_polygons_nodup <- filtered_plant_polygons %>%
  rename(SOORTSCODE = SOORTSC) %>%       # Rename first
  group_by(SOORTSCODE) %>%
  filter(!duplicated(st_as_binary(geometry)))

```

# 5. Spatial Join: Species Points with Plant Polygons and remove points in plant_polygons

```{r spatial-join}

library(sf)

# 1) compute the intersections: for each point, which polygon indices it intersects?
ints <- st_intersects(filtered_species_points, plant_polygons_nodup)

# 2) build a logical keep‐vector: TRUE if
#    • the point hits no polygons at all, OR
#    • none of the intersected polygons has the same SOORTSCODE
keep <- mapply(function(pt_code, poly_i){
  if (length(poly_i) == 0) {
    # point falls in no polygon → keep it
    return(TRUE)
  }
  # get the polygon codes for the ones it does hit
  poly_codes <- plant_polygons_nodup$SOORTSCODE[poly_i]
  # keep only if *none* match
  !any(poly_codes == pt_code, na.rm = TRUE)
},
filtered_species_points$SOORTSCODE,
ints)

# 3) subset your original sf object
species_points_cleaned <- filtered_species_points[keep, ]

```

# 6. Aggregate Species Points by Year; hier gaan in essentie kijken waar er gekarteerd is. 

```{r aggregate-points}
filtered_species_points <- filtered_species_points %>%
  st_set_geometry("geom")

# 2) buffer by 20m and union per JAAR, keeping only the original columns
species_points_buffered <- filtered_species_points %>%
  group_by(JAAR) %>%
  summarise(
    # copy the first value of every attribute except the geometry
    across(.cols = -geom, .fns = first),
    # now buffer the true geometry and union it
    geom = st_union(st_buffer(geom, 20)),
    .groups = "drop"
  )

```

# 7. Buffer Resulting Polygons (1.5 meters)

```{r buffer-polygons}
buffered_polygons <- st_buffer(species_points_buffered, dist = 1.5)
```

# 8. Union Polygons and Summarize YEARs

```{r union-polygons}
unioned_polygons <- buffered_polygons %>%
  st_union() %>%
  st_cast("POLYGON")

# Dummy YEAR_List column (in real use: summarize intersected YEARs)
unioned_polygons_sf <- st_sf(geometry = unioned_polygons) %>%
  mutate(YEAR_List = NA_character_)
```

# 9. Most Common Year Assignment per Polygon

```{r most-common-year}
# Spatial join points to unioned polygons
points_with_union <- st_join(filtered_species_points, unioned_polygons_sf, join = st_intersects)

# Add polygon IDs
unioned_polygons_sf$poly_id <- seq_len(nrow(unioned_polygons_sf))
points_with_union$poly_id <- st_nearest_feature(filtered_species_points, unioned_polygons_sf)

# Find most common JAAR for each polygon
most_common_years <- points_with_union %>%
  st_drop_geometry() %>%
  group_by(poly_id, JAAR) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(poly_id) %>%
  slice_max(order_by = count, n = 1, with_ties = FALSE) %>%
  ungroup()

# Merge back
unioned_polygons_sf <- unioned_polygons_sf %>%
  left_join(most_common_years, by = "poly_id") %>%
  rename(Most_Common_JAAR = JAAR)
```

# 10. Spatial Join Points and Polygons; Remove Mismatches

```{r filter-by-most-common-year}
points_with_most_common <- st_join(filtered_species_points, unioned_polygons_sf)

final_points <- points_with_most_common %>%
  filter(is.na(Most_Common_JAAR) | JAAR == Most_Common_JAAR)
```

# 11. Kernel Density Estimation for Each Species: (duurt erg lang, warnings zijn voor punten met identieke coordinaten. Gezien dat hier geen foute data is mag je deze warnings negeren)

```{r kernel-density}
library(sf)
library(terra)
library(spatstat)

# Get unique species codes
species_list <- unique(final_points$SOORTSCODE)

# Initialize empty list to store KDE rasters
kde_results <- list()
species_list_valid = list()


# Iterate over each species
for (species in species_list) {
  
  species_pts <- final_points %>%
    filter(SOORTSCODE == species, !is.na(Mediaan))|>
    st_transform(crs(chm_ras5))
  
  if (nrow(species_pts) >= 30) {
    species_list_valid <- c(species_list_valid, species)
    coords <- st_coordinates(species_pts)
    weights <- species_pts$Mediaan  # use Mediaan field for weights

    # Extract extent and resolution from chm_ras5
    ext_ras <- ext(chm_ras5)
    res_ras <- res(chm_ras5)   # cell size (should be 5, 5)
    
    # Define window based on chm_ras5 extent
    window <- owin(
      xrange = c(ext_ras[1], ext_ras[2]),
      yrange = c(ext_ras[3], ext_ras[4])
    )
    
    # Create ppp object using *that exact window*
    ppp_obj <- ppp(x = coords[,1], y = coords[,2], window = window, marks = weights)

    # Create an im object with exact chm_ras5 grid
    xrange <- c(ext_ras[1], ext_ras[2])
    yrange <- c(ext_ras[3], ext_ras[4])
    nx <- ncol(chm_ras5)
    ny <- nrow(chm_ras5)
    
    # Create the pixel grid for KDE
    template_im <- im(
      matrix(0, nrow = ny, ncol = nx),
      xcol = seq(xrange[1] + res_ras[1]/2, xrange[2] - res_ras[1]/2, length.out = nx),
      yrow = seq(yrange[1] + res_ras[2]/2, yrange[2] - res_ras[2]/2, length.out = ny)
    )

    
    # Run KDE using that exact pixel grid
    dens <- density(
      ppp_obj,
      weights = marks(ppp_obj),
      sigma = 2,
      at = "pixels",
      edge = TRUE,
      dimyx = c(ny, nx),
      xy = list(x = template_im$xcol, y = template_im$yrow)
    )



    # Rescale to match total number of individuals
    total_individuals <- sum(weights, na.rm = TRUE)
    pixel_sum <- sum(dens$v, na.rm = TRUE)

    if (pixel_sum > 0) {
      dens$v <- dens$v * (total_individuals / pixel_sum)
    }

    # Convert to terra raster and store
    kde_results[[species]] <- rast(dens)
    
    # Print progress
    cat("Processed species:", species, "-", nrow(species_pts), "points\n")
    
  } else {
    cat("Skipped species:", species, "- not enough points (", nrow(species_pts), ")\n")
  }
}

```

# 12. Rasterize Polygons per Species

```{r rasterize-polygons}
reference_raster <- rast(reference_raster)  # ensure it's a SpatRaster

polygon_rasters <- map(species_list_valid, function(species) {
  species_poly <- plant_polygons_nodup %>% filter(SOORTSCODE == species)
  
  if (nrow(species_poly) > 0) {
    rast_out <- rasterize(
      vect(species_poly),
      reference_raster,
      field = "abndn_2",
      fun = "mean"
    )

    # Multiply raster values (abundance per vierkante meter naar abundance per 5x5m)
    rast_out <- rast_out * 25

    rast_out
  } else {
    NULL
  }
})

```

# 13. Combine Point and Polygon Rasters per Species and write file

```{r combine-rasters}

combined_rasters <- map(species_list_valid, function(species) {
  pt <- kde_results[[species]]
  poly <- polygon_rasters[[species]]
  if (is.null(pt)) {
    return(NULL)
  }
  else{

    if (!is.null(pt)) pt <- ifel(is.na(pt), 0, pt)
    if (!is.null(poly)) poly <- ifel(is.na(poly), 0, poly)
  
    if (!is.null(pt) & !is.null(poly)) {
      raster_layer <- pt + poly
    } else if (!is.null(pt)) {
      raster_layer <- pt
    } else if (!is.null(poly)) {
      raster_layer <- poly
    } else {
      raster_layer <- NULL
    }
  
    if (!is.null(raster_layer)) {
      writeRaster(raster_layer,
                  filename = file.path(output_dir, paste0(species, "_combined.tif")),
                  overwrite = TRUE)
    }}

  raster_layer
})
names(combined_rasters) <- species_list_valid

```

# 14. Plot an Example Raster

```{r plot-example}
species_name <- names(compact(combined_rasters))[2]
plot(first_valid, main = paste("Combined raster for", species_name))

plot(kde_results[[species_name]])
```

