---
title: "vegetatie_analysis"
author: "Yolan Bosteels"
date: "2025-05-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Eerst laden we de benodigde packages en de vegetatiekaart in. We schrijven voor elk vegetatietype een raster (geotiff) uit. Dit raster is een binair raster dat aangeeft of een vegetatietype aanwezig is in een cel of niet. 

```{r cars}

library(terra)
library(googledrive)

# --- folders ---
vegetation_folder <- "C:/Users/yolan/OneDrive/Documenten/UGENT/Master/Stage/R/vegetatieklasses"

grazing_folder <- "C:/Users/yolan/OneDrive/Documenten/UGENT/Master/Stage/hoogtemetrieken"




# 0. download the .gpkg from Drive to a local file
drv_id <- as_id("1sqQMS6AcEkRAA9wgaqhI-FtdsCcwzSDF")
local_gpkg <- "vegetatiekaart.gpkg"
drive_download(
  file      = drv_id,
  path      = local_gpkg,
  overwrite = TRUE
)

# 1. now load it with terra
veg_vect <- vect(local_gpkg)


# --- 2. load a raster as template (no extra “)”) ---
template <- rast("C:/Users/yolan/OneDrive/Documenten/UGENT/Master/Stage/hoogtemetrieken/grazing_diff_norm_period_04_08.tif")

# --- 3. rasterize each Klasse to match grazing raster ---
classes <- unique(veg_vect$Klasse)

for (cls in classes) {
  # subset polygons
  sub_vect <- veg_vect[veg_vect$Klasse == cls, ]
  
  # 3a) rasterize with fun="count" → counts of polygons per cell (in principe altijd = 1 want polygonen overlappen niet)
  r_count <- rasterize(
    x          = sub_vect,
    y          = template,
    fun        = "count",
    background = 0
  )
  
  # 3b) binarize to 0/1 if desired:
  r_cls <- classify(
    r_count,
    matrix(c(
      0,   0,   0,
      1, Inf, 1
    ), ncol=3, byrow=TRUE)
  )
  # —OR— to keep raw counts, just use r_count:
  # r_cls <- r_count

  # safe filename
  cls_safe <- gsub("[^A-Za-z0-9_]", "_", cls)
  out_fname <- paste0(cls_safe, ".tif")
  
  # write GeoTIFF into grazing_folder
  writeRaster(
    r_cls,
    filename  = file.path(vegetation_folder, out_fname),
    filetype  = "GTiff",
    datatype  = "INT2U",  # 0/1 unsigned integer if binarized (or counts up to 65535)
    overwrite = TRUE
  )
  
  message("Written: ", file.path(vegetation_folder, out_fname))
}


```

Hier gebruiken we min of meer hetzelfde model als model en modelanalyse om op basis van ruimtelijke autocorrelatie en vegetatietype de begrazingsdruk te voorspellen. Het deel dat niet door deze componenten kan voorspeld worden is ons residu. Dit slaan we op. Door in een volgende stap dit residu te gaan correleren aan het voorkomen van individuele soorten krijgen we een beeld van de onafhankelijke invloed van elke individuele soort, althans dat is de hypothese. 


```{r pressure, echo=FALSE}
library(terra)
library(INLA)
library(data.table)
library(spdep)    # for knearneigh, knn2nb, nb2mat
library(MASS)    # for boxcox

#──────────────────────────────────────────────────────────────────────────────
# Loop over all grazing rasters, two‐part INLA, then residual of the sum
#──────────────────────────────────────────────────────────────────────────────
crss = " +proj=lcc +lat_0=90 +lon_0=4.36748666666667 +lat_1=49.8333339 +lat_2=51.1666672333333 +x_0=150000.01256 +y_0=5400088.4378 +ellps=intl +units=m +no_defs  "

# 0. folders + CRS
grazing_folder <- "C:/Users/yolan/OneDrive/Documenten/UGENT/Master/Stage/hoogtemetrieken"
veg_ras_folder <- "C:/Users/yolan/OneDrive/Documenten/UGENT/Master/Stage/R/vegetatieklasses"
target_crs     <- crss    # your CRS object


# 1. load classe predictors
cls_files   <- c(list.files(veg_ras_folder, pattern="\\.tif$", full.names=TRUE))
cls_names   <- sub("\\.tif$","", basename(cls_files))
class_stack <- rast(cls_files)
names(class_stack) <- cls_names

# 2. list grazing rasters
g_files <- c(
  #apr     = file.path(grazing_folder, "grazing_diff_norm_month_04.tif"),
  #mei     = file.path(grazing_folder, "grazing_diff_norm_month_05.tif"),
  #jun     = file.path(grazing_folder, "grazing_diff_norm_month_06.tif"),
  #jul     = file.path(grazing_folder, "grazing_diff_norm_month_07.tif"),
  #aug     = file.path(grazing_folder, "grazing_diff_norm_month_08.tif"),
  all_year     = file.path(grazing_folder, "grazing_density_5m")
)


# 3. process each grazing metric
for(gname in names(g_files)) {
  message(">>> ", gname)

  # 3a) load & aggregate to 25 m  
  ras <- rast(g_files[gname])
  crs(ras) <- target_crs

  # 3b) binary presence/absence raster  
  ras_bin <- ifel(ras > 0, 1, NA)  

  # 3c) estimate Box–Cox λ on the positives  
  vals_pos <- values(ras, na.rm=TRUE)
  vals_pos <- vals_pos[vals_pos > 0]
  if(length(vals_pos) < 20) {
    warning("  too few positives—skipping ", gname); next
  }
  bc     <- boxcox(vals_pos ~ 1, plotit=FALSE)
  lambda <- bc$x[which.max(bc$y)]
  message(sprintf("  λ = %.3f", lambda))

  # 3d) extract into a data.table  
  stk <- c(class_stack, ras_bin, ras)
  names(stk) <- c(cls_names, "y_bin", "y_cont")
  dt  <- as.data.table(values(stk, dataframe=TRUE, na.rm=FALSE))
  xy  <- as.data.table(xyFromCell(stk, seq_len(ncell(stk))))
  dt  <- cbind(xy, dt)[!is.na(y_bin)]
  setnames(dt, "y_cont", "obs")

  # —— NEW: drop any class predictors with no occurrences —— 
  present_cls <- cls_names[sapply(cls_names, function(cl) any(dt[[cl]] == 1, na.rm=TRUE))]
  if(length(present_cls) == 0) {
    warning("  no classes present in ", gname, " — skipping model")
    next
  }
  message("  using classes: ", paste(present_cls, collapse=", "))

  # 3e) build neighbour graph for Besag  
  coords <- as.matrix(dt[, .(x,y)])
  nb     <- knearneigh(coords, k=4) |> knn2nb(sym=TRUE)
  W      <- as(nb2mat(nb, style="B"), "sparseMatrix")
  diag(W) <- 0
  dt[, spatial_id := .I]
  g     <- inla.read.graph(W)

  # 3f) fit binary logistic  
  #    only include `present_cls` in the formula
  bin_formula <- as.formula(
    paste0("y_bin ~ 1 + f(spatial_id, model='besag', graph=g)",
           " + ", paste(present_cls, collapse=" + "))
  )
  mb <- inla(
    formula         = bin_formula,
    data            = dt,
    family          = "binomial",
    control.compute = list(dic=TRUE, waic=TRUE)
  )
  dt[, fitted_bin := mb$summary.fitted.values[,"mean"]]

  # 3g) fit positive Gaussian on transformed  
  dtp <- dt[obs > 0]
  if(nrow(dtp)==0) {
    warning("  no positives—skipping positive model")
    dtp[, fitted_pos_orig := 0]
  } else {
    if(abs(lambda) < 1e-6) {
      dtp[, y_tr := log(obs)]
    } else {
      dtp[, y_tr := (obs^lambda - 1)/lambda]
    }
    pos_formula <- as.formula(
      paste0("y_tr ~ 1 + f(spatial_id, model='besag', graph=g)",
             " + ", paste(present_cls, collapse=" + "))
    )
    mp <- inla(
      formula          = pos_formula,
      data             = dtp,
      family           = "gaussian",
      control.family   = list(link="identity"),
      control.compute  = list(dic=TRUE, waic=TRUE)
    )
    dtp[, fitted_tr := mp$summary.fitted.values[,"mean"]]

    # back‐transform to original grazing units
    if(abs(lambda) < 1e-6) {
      dtp[, fitted_pos_orig := exp(fitted_tr)]
    } else {
      dtp[, fitted_pos_orig := (fitted_tr * lambda + 1)^(1/lambda)]
    }
  }

  # 3h) merge back so everyone has fitted_pos_orig (zero where absent)
  dt[, fitted_pos_orig := 0]
  dt[dtp, on="spatial_id", fitted_pos_orig := fitted_pos_orig]

  # 3i) sum the two fitted parts and compute residual  
  dt[, fitted_sum := fitted_bin + fitted_pos_orig]
  dt[, resid_sum  := obs - fitted_sum]

  # 3j) rasterize resid_sum back to the grazing grid  
  pts   <- vect(dt[, .(x,y,resid_sum)], geom=c("x","y"), crs=crs(ras))
  r_res <- rasterize(pts, ras, field="resid_sum", background=NA)

  # 3k) write out  
  out_fn <- file.path(grazing_folder, sprintf("residual_sum_%s.tif", gname))
  writeRaster(r_res, filename=out_fn, filetype="GTiff",
              datatype="FLT4S", overwrite=TRUE)
  message("  → wrote ", out_fn, "\n")
}

message("All done!") 

```

